### 2025-06-27 paper/quarto_aaai/paper.tex
Adjust framing of "explainability"

To our surprise, the second challenge has not yet attracted major
research interest. In particular, there has been no concerted effort
towards improving the degree to which learned representations correspond
to explanations that are \textbf{interpretable} and deemed
\textbf{plausible} by humans, which we simply term ``explainability'' in
this manuscript (see Def. \ref{def-explainability}). Instead, the choice
has generally been to improve the ability of XAI tools to identify the
subset of explanations that are both plausible and valid for any given
model, independent of whether these explanations are compatible with the
learned representations \citep{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explainability can arise as a
consequence of regularization techniques aimed at other training
objectives such as generative capacity, generalization, or robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
Our work consolidates these findings within a single framework.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explainability
of models. In high-level terms, we define this concept as as the extent
to which valid explanations derived for an opaque model are also deemed
plausible with respect to the underlying data and the global
actionability constraints. To our knowledge, our framework represents
the first attempt to address this challenge by employing counterfactual
explanations already during the training phase.

@@

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:

>>>>

To our surprise, the second challenge has not yet attracted major
research interest. In particular, there has been no concerted effort
towards improving the degree to which learned representations promote
explanations that are both \textbf{interpretable} to and deemed
\textbf{plausible} by humans. Instead, the typical choice has been to
improve the ability of XAI tools to identify the subset of explanations
that are plausible and valid for any given model, independent of whether
these explanations are compatible with the learned representations
\citep{altmeyer2024faithful}. Fortunately, recent findings indicate that
improved ``explainability'' of a model can arise as a consequence of
regularization techniques aimed at other training objectives such as
generative capacity, generalization, or robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
Our contribution consolidates these findings within a unified framework.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explainability
of models. In high-level terms, we posit that the explainability of a
model depends on the extent to which valid explanations derived for this
model are also deemed plausible with respect to the underlying data and
the global actionability constraints (see Def.
\ref{def-explainability}). To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

@@

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:

### 2025-06-27 paper/sections/intro.qmd
Adjust framing of "explainability"

To our surprise, the second challenge has not yet attracted major research interest. In particular, there has been no concerted effort towards improving the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans, which we simply term "explainability" in this manuscript (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explainability can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating]. Our work consolidates these findings within a single framework.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models. In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints. To our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already during the training phase.

>>>>

To our surprise, the second challenge has not yet attracted major research interest. In particular, there has been no concerted effort towards improving the degree to which learned representations promote explanations that are both **interpretable** to and deemed **plausible** by humans. Instead, the typical choice has been to improve the ability of XAI tools to identify the subset of explanations that are plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved "explainability" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating]. Our contribution consolidates these findings within a unified framework.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models. In high-level terms, we posit that the explainability of a model depends on the extent to which valid explanations derived for this model are also deemed plausible with respect to the underlying data and the global actionability constraints (see Def. \ref{def-explainability}). To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.
